---
title: "Metabolic analysis using Compass"
author: "Putri Ramadani"
date: "2026-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Compass on dendritic cells cluster

## 1. Prepare input (HPC)
- the original compass use Scanpy and other Python-based toolsfor pseudobulking, but here I use AggregateExpression in Seurat to do pseudobulking. 
- The key is to make sure the output files are in the format that Compass expects (expression.mtx, genes.tsv, sample_names.tsv, and matedata).
```{r}
# load Seurat object 
load("/cluster/projects/aokigroup/t138390uhn/scRNAseq_ALCL_FFPE/data/Seurat_objects/subclustering_2603/Myeloid_subclustering_res0.2_annotated.Rdata")

# check column names of metadata
colnames(myeloid@meta.data)

table(myeloid$myeloid_celltype, myeloid$StudyID)

DCs <- subset(myeloid, subset = myeloid_celltype == "cDC1" | myeloid_celltype == "cDC2" | myeloid_celltype == "pDC")

ncol(DCs) # [1] 3274
nrow(DCs) # [1] 18518

# dir
out_dir <- "/cluster/home/t138390uhn/scRNAseq_ALCL_FFPE/data/processed/compass_DCs"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

DefaultAssay(DCs) <- "RNA"

# use a "--" separator 
DCs$pbulk_group <- paste(DCs$myeloid_celltype, DCs$StudyID, sep = "--")

# --- 1) Pseudobulk ---
pbulk <- AggregateExpression(
  DCs, 
  assays   = "RNA",
  slot     = "counts",
  group.by = "pbulk_group",
  return.seurat = FALSE
)

pbulk_counts <- pbulk$RNA   # genes x pseudobulks
nrow(pbulk_counts)
ncol(pbulk_counts)
# Pseudobulk matrix: 18518 genes x 120 pseudobulks

# --- 2) CPM normalize ---
lib_sizes <- colSums(pbulk_counts)
cpm <- t(t(pbulk_counts) / lib_sizes) * 1e6

# --- 3) Write Compass inputs ---
writeMM(as(cpm, "sparseMatrix"), file = file.path(out_dir, "expression.mtx"))
## inside expression.mtx:
## Rows = genes 
## Columns = pseudobulks 
## Values = CPM expression (counts per million, library-size normalized)

write.table(rownames(cpm),
            file = file.path(out_dir, "genes.tsv"),
            sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = FALSE)

write.table(colnames(cpm),
            file = file.path(out_dir, "sample_names.tsv"),
            sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = FALSE)

# note:
# expression.mtx = the numbers
# genes.tsv = the gene names (matches the rows in order)
# sample_names.tsv = the pseudobulk names (matches the columns in order)


# --- 4) Metadata ---
pbulk_meta <- data.frame(
  pbulk_id = colnames(cpm),
  stringsAsFactors = FALSE
)
pbulk_meta$myeloid_celltype <- sub("--.+$", "", pbulk_meta$pbulk_id)
pbulk_meta$StudyID          <- sub("^.+--", "", pbulk_meta$pbulk_id)

group_sizes <- table(DCs$pbulk_group)
pbulk_meta$n_cells <- as.integer(group_sizes[pbulk_meta$pbulk_id])

patient_info <- unique(
  DCs@meta.data[, c("StudyID", "Disease", "Relapse", "Subtype", "Category1", "Category2", "SPN_status", "CIRBP_status", "LGALS3_status")]
)
pbulk_meta <- dplyr::left_join(pbulk_meta, patient_info, by = "StudyID")

write.table(pbulk_meta,
            file = file.path(out_dir, "pbulk_metadata.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

# check output:
print(list.files(out_dir))
print(range(colSums(cpm)))  
print(head(pbulk_meta))
# [1] "expression.mtx"     "genes.tsv"          "pbulk_metadata.tsv"
# [4] "sample_names.tsv"  
# [1] 1e+06 1e+06

       pbulk_id myeloid_celltype StudyID n_cells  Disease    Relapse Subtype
1 cDC1--ALCL001             cDC1 ALCL001      15 ALK+ALCL No Relapse  Type I
2 cDC1--ALCL002             cDC1 ALCL002       3 ALK+ALCL No Relapse  Type I
3 cDC1--ALCL003             cDC1 ALCL003      17 ALK+ALCL    Relapse  Type I
4 cDC1--ALCL004             cDC1 ALCL004       1 ALK+ALCL No Relapse  Type I
5 cDC1--ALCL005             cDC1 ALCL005       2 ALK+ALCL No Relapse  Type I
6 cDC1--ALCL006             cDC1 ALCL006      24 ALK+ALCL No Relapse  Type I
  Category1      Category2 SPN_status CIRBP_status LGALS3_status
1 Pediatric pediatric_ALCL   negative     positive      positive
2 Pediatric pediatric_ALCL   negative     negative      positive
3 Pediatric pediatric_ALCL   negative     positive      negative
4 Pediatric pediatric_ALCL   negative     negative      positive
5 Pediatric pediatric_ALCL   negative     negative      positive
6 Pediatric pediatric_ALCL   negative     positive      positive

```


## 2. Perform Compass
- download new gurobi license
-  perform Compass on Nibi (Canada Alliance) because Compass and Gurobi need internet

### Move outputs to Nibi Alliance
```{bash}
scp -r /Users/putriramadani/Documents/GitHub/scRNAseq_ALCL_FFPE/data/processed/compass_DCs/ taoki@nibi.alliancecan.ca:~/
```

### Nibi on login node 
note: make sure to request access to nibi through Tomo's account, and prepare gurobi license in the same dir as compass_DCs.sh
compass_DCs.sh
```{bash}
#!/bin/bash
#SBATCH --account=def-taoki_cpu
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=32
#SBATCH --mem=64G
#SBATCH --time=5:00:00
#SBATCH --job-name=compass_DCs
#SBATCH --output=%x-%j.out
#SBATCH --error=%x-%j.err

unset PYTHONPATH
source ~/miniconda3/bin/activate compass

cd ~/compass_DCs
compass --data-mtx expression.mtx genes.tsv sample_names.tsv \
        --num-processes 32 \
        --species homo_sapiens

```


## 3. Analysis (local RStudio)
- Compass result analysis is available on https://compass-wagnerlab.readthedocs.io/en/latest/ 
- Here we generate a heatmap and rank the pathway enrichment result combining Compass result (reactions.tsv) and GSEA
```{r}
library(fgsea)
library(pheatmap)
library(viridisLite)
library(tidyr)

# ---- subset names to compare ----
subsets_to_use <- c("cDC1", "cDC2", "pDC")


# ---- Load Compass output ----
# Compass outputs "penalties" per reaction: lower penalty = more consistent
# (more feasible) flux through that reaction given the expression data.
reactions <- read.table("DCs/reactions.tsv", sep = "\t", header = TRUE,
                        row.names = 1, check.names = FALSE)

# Standard Compass transform: flip penalties into "consistency" scores so that
# HIGHER values = more metabolically active/consistent reaction (easier to interpret and matches the convention used in Compass tutorials/papers).
react_mat <- -log(as.matrix(reactions) + 1)


# ---- Load metadata, filter to subsets of interest ----
pbulk_meta <- read.table("DCs/pbulk_metadata.tsv", sep = "\t", header = TRUE,
                         stringsAsFactors = FALSE)
meta_use <- pbulk_meta[pbulk_meta$myeloid_celltype %in% subsets_to_use, ]

# Keep only pseudobulk samples that exist in BOTH the reaction matrix and
# the filtered metadata (keeps the two objects aligned to the same samples)
react_use <- react_mat[, colnames(react_mat) %in% meta_use$pbulk_id]
meta_use <- meta_use[meta_use$pbulk_id %in% colnames(react_use), ]

print(table(meta_use$myeloid_celltype))
# cDC1 cDC2  pDC 
# 42   45   33


# ---- Load Recon2 metadata and filter reactions ----
reaction_metadata <- read.csv("extdata/RECON2/reaction_metadata.csv", row.names = 1)

# Compass reaction IDs have a "_pos"/"_neg" suffix (forward/reverse direction
# of a reversible reaction) that isn't in the Recon2 metadata table, so strip
# it to look up each reaction's annotation.
base_ids <- sub("_(pos|neg)$", "", rownames(react_use))
ms <- reaction_metadata[base_ids, ]

# CHECK: if any base_ids aren't found in reaction_metadata, R silently
# returns an all-NA row instead of erroring, and those reactions get quietly
# dropped by the `keep` filter below. Worth confirming this is 0 (or small and expected) 
n_unmatched <- sum(!base_ids %in% rownames(reaction_metadata))
print(n_unmatched) 
# 0

# Keep only reactions with:
#  - confidence 0 (no confidence penalty applied) or 4 (highest confidence)
#  - a known EC number (i.e. an actual annotated enzyme, not a generic/artificial reaction)
keep <- ms$confidence %in% c(0, 4) & !is.na(ms$EC_number)

# Citric acid cycle reactions that are NOT in the mitochondrial compartment
# ([m]) are usually cytosolic/duplicated versions of the same reaction --
# recode these to "Other" so the "Citric acid cycle" pathway set only contains the true mitochondrial TCA reactions.
not_mito_tca <- !grepl("\\[m\\]", ms$formula) & ms$subsystem == "Citric acid cycle"
ms$subsystem[not_mito_tca] <- "Other"

react_use <- react_use[keep, ]
ms <- ms[keep, ]


# ---- Build subsystem -> reaction list (gene sets for GSEA) ----
reaction_subsys <- setNames(ms$subsystem, rownames(react_use))
pathways <- split(names(reaction_subsys), reaction_subsys)

# Drop subsystems that aren't real metabolic pathways (transport steps,
# unassigned/misc buckets, exchange reactions) -- these aren't meaningful
# units for pathway enrichment.
exclude <- c("Transport, extracellular", "Transport, mitochondrial",
             "Transport, peroxisomal", "Transport, lysosomal",
             "Transport, golgi apparatus", "Transport, endoplasmic reticular",
             "Transport, nuclear", "Unassigned", "Miscellaneous",
             "Exchange/demand reaction", "Other")
pathways <- pathways[!names(pathways) %in% exclude]

# Keep pathway sizes GSEA can reasonably test (too small = unstable, too large = uninformative/non-specific)
pathways <- pathways[lengths(pathways) >= 5 & lengths(pathways) <= 500]


# ---- Drop zero-variance reactions before ranking ----
# Reactions with identical (e.g. all-zero) consistency scores across every
# sample in both groups carry no information and produce a Cohen's d of
# exactly 0 or NA. Leaving them in creates a large block of tied ranks,
# which is likely the cause of the "ties in preranked stats" warning fgsea
# threw (40%+ tied). Removing them doesn't change the biology, just makes
# the ranking cleaner and the tie warning go away.
zero_var <- apply(react_use, 1, function(r) sd(r, na.rm = TRUE) == 0)
cat("Zero-variance reactions removed:", sum(zero_var), "of", length(zero_var), "\n")
react_use <- react_use[!zero_var, ]
ms <- ms[!zero_var, ]


# ---- Cohen's d effect size: compared subsets  ----
cohens_d <- function(x, y) {
  # na.rm = TRUE added: without it, a single NA in x or y makes var() return
  # NA, which silently makes the whole Cohen's d NA for that reaction (and it
  # then gets dropped later at `ranks[is.finite(ranks)]` with no warning).
  x <- x[!is.na(x)]
  y <- y[!is.na(y)]
  if (length(x) < 2 || length(y) < 2) return(NA)
  if (sd(c(x, y)) == 0) return(NA)
  pooled <- sqrt(((length(x) - 1) * var(x) + (length(y) - 1) * var(y)) /
                   (length(x) + length(y) - 2))
  (mean(x) - mean(y)) / pooled
}

# CHECK: confirm whether missingness is actually present in the data.
# if it's not 0, it explains why some reactions were disappearing from ranking.
cat("Total NAs in react_use:", sum(is.na(react_use)), "\n")
# Total NAs in react_use: 0 


# ---- For each subset: compared subsets Cohen's d -> GSEA ----
results <- list()
for (subset in subsets_to_use) {
  in_ids  <- meta_use$pbulk_id[meta_use$myeloid_celltype == subset]
  out_ids <- meta_use$pbulk_id[meta_use$myeloid_celltype != subset]
  cat("Subset:", subset, "| in:", length(in_ids), "out:", length(out_ids), "\n")
  if (length(in_ids) < 3 || length(out_ids) < 3) next
  
  # Effect size per reaction: positive = higher consistency in this subset than in the rest, negative = lower
  ranks <- vapply(seq_len(nrow(react_use)), function(i) {
    cohens_d(react_use[i, in_ids], react_use[i, out_ids])
  }, numeric(1))
  names(ranks) <- rownames(react_use)
  ranks <- ranks[is.finite(ranks)]
  
  
  set.seed(2026)
  res <- fgsea(pathways = pathways, stats = ranks,
               minSize = 5, maxSize = 500, scoreType = "pos")
  res$subset <- subset
  results[[length(results) + 1]] <- res
}
all_res <- do.call(rbind, results)
# -- output --
# Subset: cDC1 | in: 42 out: 78 
# Subset: cDC2 | in: 45 out: 75 
# Subset: pDC | in: 33 out: 87 
# Warning messages:
#  1: In preparePathwaysAndStats(pathways, stats, minSize, maxSize, gseaParam,  :
#   There are ties in the preranked stats (42.72% of the list).
#   The order of those tied genes will be arbitrary, which may produce unexpected results.
#  2: In preparePathwaysAndStats(pathways, stats, minSize, maxSize, gseaParam,  :
#   There are ties in the preranked stats (42.65% of the list).
#   The order of those tied genes will be arbitrary, which may produce unexpected results.
#  3: In preparePathwaysAndStats(pathways, stats, minSize, maxSize, gseaParam,  :
#   There are ties in the preranked stats (42.72% of the list).
#   The order of those tied genes will be arbitrary, which may produce unexpected results.


# ---- Build heatmap: pathways x subsets ----
# Cap -log10(padj) at 4 so a handful of extremely significant pathways don't wash out the color scale for everything else
all_res$score <- pmin(-log10(all_res$padj), 4)
all_res$score[!is.finite(all_res$score)] <- 0

heat_mat <- as.data.frame(
  pivot_wider(all_res[, c("pathway", "subset", "score")],
              names_from = subset, values_from = score)
)
rownames(heat_mat) <- heat_mat$pathway
heat_mat$pathway <- NULL
heat_mat <- heat_mat[, subsets_to_use]
heat_mat[is.na(heat_mat)] <- 0

# Keep pathways with a strong signal in at least one subset (score >= 2,
# i.e. padj <= 0.01) AND that aren't significant across every subset
# (score >= 1, i.e. padj <= 0.1) -- this filters for pathways that are both
# strongly and somewhat SPECIFICALLY enriched, rather than generically active everywhere
strong  <- rowSums(heat_mat >= 2) > 0
not_all <- rowSums(heat_mat >= 1) <= length(subsets_to_use) - 1
heat_mat_filt <- heat_mat[strong & not_all, ]
cat("Pathways shown:", nrow(heat_mat_filt), "\n")


# ---- Plot ----
pheatmap(
  as.matrix(heat_mat_filt),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = viridis(100),
  border_color = "black",
  fontsize_row = 10,
  fontsize_col = 11,
  main = "Metabolic pathway enrichment per DC cluster",
  filename = "260811_DCs_metabolic_pathway_heatmap.png",
  width = 8, height = 8
)

```